\bf Supervised Tensor Decomposition with Interactive Side Information

### 2020-09-08 tensor_regress/jasa-latexsource/tensor-jasa.tex
update

\bigskip
\begin{abstract}
Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. Identifying the relationship between a high-dimensional tensor and side information is important yet challenging. Here, we develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike unsupervised tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating updating algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We identify the key global connectivity pattern and pinpoints the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, the package and data used are available at~\url{https://CRAN.R-project.org/package=tensorregress}.

@@

We present a general model and associated method for decomposing a data tensor with exponential family entries and interactive side information. 
We formulate the learning task as a structured regression problem, with tensor observation serving as the response, and the multiple side information as interactive features. Figure~\ref{fig:intro1}b illustrates our model in the special case of order-3 tensors. A low-rank structure is imposed to the conditional mean tensor, where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ are assumed to belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

In addition, we use tools from generalized linear model (GLM) to allow heteroscedacity due to the mean-variance relationship in the non-Gaussian data. This flexibility is important in practice. In classical GLM, the sample size and feature dimension are well defined; however, in the tensor data analysis, we observe only one single realization of tensor response but up to $K$ interactive feature matrices, where $K$ is the number of tensor modes. Both the sample size and feature dimension grow exponentially in $K$. We use a low-rank constraint in the conditional mean tensor (with a suitable transformation) to mitigate the curse of high dimensionality. The statistical convergence of our estimator is established, and we quantify the gain in predictive power through several case studies.

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of a data tensor. In contrast, our model is a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and multiple features. The low-rank factorization is determined jointly by the tensor data and feature matrices on multiple sides.

The second line of work studies the scalar-to-tensor regression, in which the response is a scalar and the predictor is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to theirs because we treat the data tensor as random variables. The difference in model settings yields different focuses and interpretations. Consider the neuroimaging analysis as an example. The scalar-to-tensor regression aims to predict the individual response based on brain connectivity network, while tensor-to-scalar model focuses on understanding the variation in the brain connectivity attributable to the individual features such as age, gender, and disease status. The scalar-to-tensor model is insufficient for tensor dimension reduction because it lacks the modeling of stochastic noise in the data tensor.

The third line of work studies the tensor-on-tensor regression, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, gahrooei2020multiple}. Our model shares a common ground but provides a more efficient context-specific solutions than earlier appraoches. As we show in Section~\ref{subsec:statprob}, the supervised tensor decomposition 
has an interesting interpretation as a special tensor-on-tensor regression. Nevertheless, our work improves from previous work in scope and applicability. Previous methods~\citep{gahrooei2020multiple,lock2018tensor} mainly focus on Gaussian tensor. The Frobenius norm used in the objective is statistical suboptimal for general exponential family tensors. Maximum likelihood estimator (MLE) is studied in~\cite{raskutti2015convex} and a convex relaxation algorithm is proposed to solve for low-rank tensor coefficients. However, in the tensor case, convex MLE suffers from both computational intractability and statistical suboptimality. We advocate a non-convex approach and provide strong evidence for its success in our setting. Most previous tensor regression focuses on prediction~\citep{lock2018tensor,raskutti2015convex,gahrooei2020multiple}, and we go step further by finding the sufficient dimension reduction~\citep{adragni2009sufficient}, $\text{Span}(\mM_k)$, that facilitates the identification of effective features in prediction (see Figure~\ref{fig:intro1}b). The latter approach greatly improves the \emph{interpretability} in prediction. In this regards, our method opens up new opportunities for tensor data analysis in a wider range of possible applications.

@@

We introduce the basic tensor properties used in the paper; more details on tensor notation can be found in~\citet{kolda2009tensor}. We use $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ to denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor, where $K$ corresponds to the number of modes of $\tY$ and is called the order. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as
\begin{equation}
\tY \times_1\mX_1\ldots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\ldots x_{j_K,i_K}^{(K)}},
\end{equation}
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use the shorthand $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as 
\[
\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.
\] 
The Frobenius norm and maximum norm of $\tY$ are defined as 
\[
\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}\quad \text{and}\quad \mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.
\]
 A higher-order tensor can be reshaped into a lower-order object. We use $\text{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ the operation that reshapes the tensor along mode-$k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The multilinear rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\text{Unfold}_k(\tY)$, $k=1,\ldots,K$. Given a matrix $\mM\in\mathbb{R}^{d\times r}$, we use $\text{Span}(\mM)$ to denote the space spanned by columns of $\mM$. We use lower-case letters (e.g.,\ $a,\ b,\ c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA,\ \mB,\ \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA,\ \tB,\  \tC$) for tensors of order three or greater. We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$. For ease of notation, we allow the basic arithmetic operators (e.g.,\ $+$, $-$, $\geq $) and univariate 
functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner.

\section{Motivation and model}\label{sec:model}
\subsection{General framework for supervised tensor decomposition}
We begin with a general framework for supervised tensor decomposition and then discuss its implication in three concrete examples. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We propose a multilinear conditional mean model between the data tensor and feature matrices. Assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent random variables following an exponential family distribution, and the conditional mean admits the form
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\quad \text{with}\quad \Theta =\tB\times\{\mX_1,\ \ldots,\ \mX_K\} ,
\end{align}
where $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the unknown parameter tensor, $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$, and $\times$ denotes the tensor-by-matrix product. The choice of link function is based on the assumed distribution family of tensor entries. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution. In general, dispersion parameters can also be included in the model. Because our main focus is the tensor decomposition under the mean model, we omit the dispersion parameter in this section for ease of presentation.

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1,\ \ldots,\ \mX_K\mM_K\}\right),\notag\\
\text{where} \ &\ \mM_k\mM^T_k=\mI_{p_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\ \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. The model formulation~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$; the adaptation to unknown rank will be described in Section~\ref{sec:tuning}. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. Note that the features $\mX_k$ affect the distribution of tensor $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes. Our goal is to find $\mM_k$ and the corresponding $\tC$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

\subsection{Three examples}
We give three concrete examples of supervised tensor decomposition model~\eqref{eq:decomp} that arises in practice. 
\begin{example}[Spatio-temporal growth model]
Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth model can be represented as
\begin{equation}\label{eq:time}
\mathbb{E}(\tY|\mX_1,\mX_2,\mX_3)=\tC\times\{\mX_1\mM_1,\ \mX_2\mM_2,\ \mX_3\mM_3\},
\end{equation}
where $\mX_1=\text{blockdiag}\{\mathbf{1}_p,\ldots,\mathbf{1}_p\}\in \{0,1\}^{d\times q}$ is the design matrix for lake types, and
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\quad
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively, $\tC\in\mathbb{R}^{r_1\times r_2\times r_3}$ is the unknown core tensor, and $\mM_k$ are unknown dimension reduction matrices of coherent sizes. The model~\eqref{eq:time} is a higher-order extension of the bilinear ``growth curve'' model originally proposed for matrix data~\citep{gabriel1998generalised,potthoff1964generalized,srivastava2008models}. The factors $\mX_k\mM_k$ are sufficient features for the mean model~\eqref{eq:time}. Clearly, the spatial-temporal model is a special case of our supervised tensor decomposition model, with features available on each of the three modes.
\end{example}

\begin{example}[Network population model] 
Network response model is recently developed in the context of neuroimanig analysis. The goal is to study the relationship between brain network connectivity pattern and the individual features. Suppose we have a sample of $n$ observations, $\{(\mY_i, \mx_i)\colon i=1,\ldots,n\}$, where for each individual $i\in[n]$, $\mY_i\in\{0,1\}^{d\times d}$ is the symmetric adjacency matrix whose entries indicate presences/absences of connectivities between $d$ brain nodes, and $\mx_i\in\mathbb{R}^p$ is the individual's feature such as age, gender, cognition score, etc. The network-response model~\citep{rabusseau2016low, zhang2018network} has the conditional mean
\begin{equation}\label{eq:network}
\text{logit}(\mathbb{E}(\mY_i|\mx_i))=\tB\times_3\mx_i, \quad \text{for }i=1,\ldots,n,
\end{equation}
where $\tB\in \mathbb{R}^{d\times d\times p}$ is a coefficient tensor of rank $r(\tB)=(r,r,r')$.

@@

 \begin{example}[Dyadic data with node attributes] Dyadic dataset consists of measurements on pairs of objects or under a pair of conditions. Common examples include graphs and networks. Let $\tG=(V,E)$ denote a graph, where $V=[d]$ is the node set of the graph, and $E\subset V\times V$ is the edge set. Suppose that we also observe feature vector $\mx_i\in\mathbb{R}^p$ associated to each $i\in V$. A probabilistic model on the graph $\tG=(V,E)$ can be described by the following matrix regression. The edge connects the two vertices $i$ and $j$ independently of other pairs, and the probability of connection is modeled as
\begin{equation}\label{eq:edge}
 \text{logit}\left(\mathbb{P}\left((i,j)\in E\right)\right)=\mx^T_i\mB\mx_j=\langle \mB,\ \mx^T_i\mx_j\rangle,
 \end{equation}
 where $\mB\in\mathbb{R}^{p\times p}$ is a symmetric rank-$r$ matrix. The low-rankness in $\mB$ has demonstrated its success in modeling transitivity, balance, and communities in the networks~\citep{hoff2005bilinear}. We show that our supervised tensor decompostiion~\eqref{eq:tensormodel} also incorporates the graph model as a special case. Let $\tY=\entry{y_{ij}}$ be a binary matrix where $y_{ij}=\mathds{1}_{(i,j)\in E}$. Define $\mX=[\mx_1,\ldots,\mx_n]^T\in\mathbb{R}^{n\times p}$. Then, the graph model~\eqref{eq:edge} can be expressed as
 \[
 \text{logit}(\mathbb{E}(\mY|\mX))=\mC\times\{\mX\mM,\ \mX\mM\}, 
  \]
  where $\mC\in\mathbb{R}^{r\times r}$ and $\mM\in\mathbb{R}^{p\times r}$ can be taken as the singular value decomposition of $\mB=\mM\mC\mM^T$. 
  \end{example}

>>>>

\bigskip
\begin{abstract}
Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. Identifying the relationship between a high-dimensional tensor and side information is important yet challenging. Here, we develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike unsupervised tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We identify the key global connectivity pattern and pinpoints the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, the package and data used are available at~\url{https://CRAN.R-project.org/package=tensorregress}.

@@

We present a general model and associated method for decomposing a data tensor with exponential family entries and interactive side information. 
We formulate the learning task as a structured regression problem, with tensor observation serving as the response, and the multiple side information as interactive features. Figure~\ref{fig:intro1}b illustrates our model in the special case of order-3 tensors. A low-rank structure is imposed to the conditional mean tensor, where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

In addition, we use tools from generalized linear model (GLM) to allow heteroscedacity due to the mean-variance relationship in the non-Gaussian data. This flexibility is important in practice. In classical GLM, the sample size and feature dimension are well defined; however, in the tensor data analysis, we observe only one single realization of tensor response but up to $K$ interactive feature matrices, where $K$ is the number of tensor modes. Both the sample size and feature dimension grow exponentially in $K$. We use a low-rank constraint in the conditional mean tensor (with a suitable transformation) to mitigate the curse of high dimensionality. The statistical convergence of our estimator is established, and we quantify the gain in prediction through several case studies.

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of a data tensor. In contrast, our model is a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and multiple features. The low-rank factorization is determined jointly by the tensor data and feature matrices.

The second line of work studies the scalar-to-tensor regression, in which the response is a scalar and the predictor is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to this line because we treat the data tensor as random variables. The difference in model settings yields different focuses and interpretations. Consider the neuroimaging analysis as an example. The scalar-to-tensor regression aims to predict the individual response based on brain connectivity network, while tensor-to-scalar model focuses on understanding the variation in the brain connectivity attributable to the individual features such as age, gender, and disease status. The scalar-to-tensor model is insufficient for tensor dimension reduction because it lacks the modeling of stochastic noise in the data tensor.

The third line of work studies the tensor-on-tensor regression, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, gahrooei2020multiple}. Our model shares a common ground but provides a more efficient context-specific solutions than earlier approaches. As we show in Section~\ref{subsec:statprob}, the supervised tensor decomposition 
has an interesting interpretation as a special tensor-on-tensor regression. Nevertheless, our work improves from previous work in scope and applicability. Previous methods~\citep{gahrooei2020multiple,lock2018tensor} mainly focus on Gaussian tensors. The Frobenius norm used in the objective function is statistical suboptimal for general exponential family tensors. Maximum likelihood estimator (MLE) is studied in~\cite{raskutti2015convex} and a convex relaxation algorithm is proposed to solve for low-rank tensor coefficients. However, in the tensor case, convex MLE suffers from both computational intractability and statistical suboptimality. We advocate a non-convex approach and provide strong evidence for its success in our setting. Most previous tensor regression focuses on prediction~\citep{lock2018tensor,raskutti2015convex,gahrooei2020multiple}, and we go step further by finding the sufficient dimension reduction~\citep{adragni2009sufficient}, $\text{Span}(\mM_k)$, that facilitates the identification of effective features in prediction (see Figure~\ref{fig:intro1}b). The latter approach greatly improves the \emph{interpretability} in prediction. In this regards, our method opens up new opportunities for tensor data analysis in a wider range of possible applications.

@@

We introduce the basic tensor properties used in the paper; more details on tensor notation can be found in~\citet{kolda2009tensor}. We use $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$\fixme{Miaoyan}{perhaps use $y_\omega$} to denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor, where $K$ corresponds to the number of modes of $\tY$ and is called the order. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as
\begin{equation}
\tY \times_1\mX_1\ldots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\ldots x_{j_K,i_K}^{(K)}},
\end{equation}
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use the shorthand $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as 
\[
\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.
\] 
The Frobenius norm and maximum norm of $\tY$ are defined as 
\[
\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}\quad \text{and}\quad \mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.
\]
 A higher-order tensor can be reshaped into a lower-order object. We use $\text{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ the operation that reshapes the tensor along mode-$k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The multilinear rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\text{Unfold}_k(\tY)$, $k=1,\ldots,K$. Given a matrix $\mM\in\mathbb{R}^{d\times r}$, we use $\text{Span}(\mM)$ to denote the space spanned by columns of $\mM$. We use lower-case letters (e.g.,\ $a,\ b,\ c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA,\ \mB,\ \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA,\ \tB,\  \tC$) for tensors of order three or greater. We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$. For ease of notation, we allow the basic arithmetic operators (e.g.,\ $+$, $-$, $\geq $) and univariate 
functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner.

\section{Motivation and model}\label{sec:model}
\subsection{General framework for supervised tensor decomposition}
We begin with a general framework for supervised tensor decomposition and then discuss its implication in three concrete examples. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We propose a multilinear conditional mean model between the data tensor and feature matrices. Assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\quad \text{with}\quad \Theta =\tB\times\{\mX_1,\ \ldots,\ \mX_K\} ,
\end{align}
where $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the unknown parameter tensor, $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$, and $\times$ denotes the tensor-by-matrix product. The choice of link function is based on the assumed distribution family of tensor entries. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution. In general, dispersion parameters can also be included in the model. Because our main focus is the tensor decomposition under the mean model, we omit the dispersion parameter in this section for ease of presentation.

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1,\ \ldots,\ \mX_K\mM_K\}\right),\notag\\
\text{where} \ &\ \mM_k\mM^T_k=\mI_{p_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\ \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. Note that model~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$; the adaptation to unknown rank will be described in Section~\ref{sec:tuning}. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes. Our goal is to find $\mM_k$ and the corresponding $\tC$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

\subsection{Three examples}
We give three concrete examples of supervised tensor decomposition model~\eqref{eq:decomp} that arises in practice. 
\begin{example}[Spatio-temporal growth model]
The growth curve model~\citep{gabriel1998generalised,potthoff1964generalized,srivastava2008models} was originally proposed as an example of bilinear model for matrix data, and we adopt its higher-order extension here. Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth model can be represented as
\begin{equation}\label{eq:time}
\mathbb{E}(\tY|\mX_1,\mX_2,\mX_3)=\tC\times\{\mX_1\mM_1,\ \mX_2\mM_2,\ \mX_3\mM_3\},
\end{equation}
where $\mX_1=\text{blockdiag}\{\mathbf{1}_p,\ldots,\mathbf{1}_p\}\in \{0,1\}^{d\times q}$ is the design matrix for lake types, and
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\quad
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively, $\tC\in\mathbb{R}^{r_1\times r_2\times r_3}$ is the unknown core tensor, and $\mM_k$ are unknown dimension reduction matrices of coherent sizes. The factors $\mX_k\mM_k$ are sufficient features for the mean model~\eqref{eq:time}. The spatial-temporal model is a special case of our supervised tensor decomposition model, with features available on each of the three modes.
\end{example}

\begin{example}[Network population model]\label{example:brain}
Network response model~\citep{rabusseau2016low, zhang2018network} is recently developed for neuroimanig analysis. The goal is to study the relationship between brain network connectivity pattern and features of individuals. Suppose we have a sample of $n$ observations, $\{(\mY_i, \mx_i)\colon i=1,\ldots,n\}$, where for each individual $i\in[n]$, $\mY_i\in\{0,1\}^{d\times d}$ is the symmetric adjacency matrix whose entries indicate presences/absences of connectivities between $d$ brain nodes, and $\mx_i\in\mathbb{R}^p$ is the individual's feature such as age, gender, cognition score, etc. The network-response model  has the conditional mean
\begin{equation}\label{eq:network}
\text{logit}(\mathbb{E}(\mY_i|\mx_i))=\tB\times_3\mx_i, \quad \text{for }i=1,\ldots,n,
\end{equation}
where $\tB\in \mathbb{R}^{d\times d\times p}$ is a coefficient tensor of rank $r(\tB)=(r,r,r')$.

@@

 \begin{example}[Dyadic data with node attributes] Dyadic dataset consists of measurements on pairs of objects or under a pair of conditions. Common examples include graphs and networks. Let $\tG=(V,E)$ denote a graph, where $V=[d]$ is the node set of the graph, and $E\subset V\times V$ is the edge set. Suppose that we also observe feature vector $\mx_i\in\mathbb{R}^p$ associated to each $i\in V$. A probabilistic model on the graph $\tG=(V,E)$ can be described by the following matrix regression. The edge connects the two vertices $i$ and $j$ independently of other pairs, and the probability of connection is modeled as
\begin{equation}\label{eq:edge}
 \text{logit}\left(\mathbb{P}\left((i,j)\in E\right)\right)=\mx^T_i\mB\mx_j=\langle \mB,\ \mx^T_i\mx_j\rangle,
 \end{equation}
 where $\mB\in\mathbb{R}^{p\times p}$ is a symmetric rank-$r$ matrix. The low-rankness in $\mB$ has demonstrated its success in modeling transitivity, balance, and communities in the networks~\citep{hoff2005bilinear}. We show that our supervised tensor decompostiion~\eqref{eq:tensormodel} also incorporates the graph model as a special case. Let $\tY=\entry{y_{ij}}$ be a binary matrix where $y_{ij}=\mathds{1}_{(i,j)\in E}$. Define $\mX=[\mx_1,\ldots,\mx_n]^T\in\mathbb{R}^{n\times p}$. Then, the graph model~\eqref{eq:edge} can be expressed as
 \[
 \text{logit}(\mathbb{E}(\mY|\mX))=\mC\times\{\mX\mM,\ \mX\mM\}, 
  \]
  where $\mC\in\mathbb{R}^{r\times r}$ and $\mM\in\mathbb{R}^{p\times r}$ are based on the singular value decomposition of $\mB=\mM\mC\mM^T$. 
  \end{example}
